---
title: 'Exercise 5: Dimensionality reduction techniques'
author: "Harri Lindroos"
date: "2 joulukuuta 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#RStudio exercise 5: Dimensionality reduction techniques

#Reading the data

This weeks data is provided by United Nations Development Programme, and it consists of different variables concerning human development. Our prepared dataset is a partial combination of two indices: The Human Development Index (HDI) and The Gender Inequality Index. As we can see from below, the dataset consists of 155 countries (observations) and 8 variables: 

```{r}
hdgii <- read.csv("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/human2.txt")
dim(hdgii)
```

The variables measure per capita gross national income (GNI), life expectancy at birth (Life.Exp), expected years of schooling (Edu.Exp), maternal mortality ratio (Mat.Mor), adolescent birth rate (Ado.Birth), Percetange of female representatives in parliament (Parli.F), the ratio of females with at least secondary education compared to men with secondary education (Edu2.FM), and the ratio of female and men populations in the labour force (Labo.FM). The values and ranges of all the variables can be observed below:

```{r}
summary(hdgii)
```

#Correlations

Next we are going to take a look at the correlations among our data. One simple and elightening way to examine this is to calculate a correlations matrix and to plot it. Our correlations plot colours negative correlations (0>x>-1) with different shades of red and positive correlations (-1>x>0) with shades of blue. Higher correlation values (the bigegr the distance is from the zero) are also presented with bigger dots:

```{r}
library(dplyr)
library(corrplot)
library(GGally)
cor(hdgii) %>% corrplot(method="circle", type="upper", cl.pos="b", tl.pos="d", tl.cex = 0.6)
```

Judging by the above plot the two stand out correlations are the positive correlation for life expectancy at birth and expected years of schooling, and the negative one for maternal mortality ratio and life expectancy at birth. In my opinion these correlations fit very well with common sense presumptions about human development: When . It is also possible to examine these correlations in more precise manner by plotting the same correlations matrix with numerical values. This matrix plot confirms our previous assumptions concerning the strongest correlations: The exact numerical value for the correlation between life expectancy at birth and expected years of schooling is 0.789 and the correlation for maternal mortality ratio and life expectancy at birth is -0.857. The later is the highest correlation in our data. The lowest correlations are the values of ~ 0.0096 for Labo.FM and Edu2.FM and ~ -0.0071 for Ado.Birth and Parli.F. These correlations are very small and the former one is nearly non-existant. All the correlations can be examined from below. 

```{r}
ggpairs(hdgii)
```

##PCA

```{r}
pca_hdgii <- prcomp(hdgii)
summary(pca_hdgii)
```

```{r}
biplot(pca_hdgii, choices = 1:2, cex = c(0.8, 1), col = c("grey40", "deeppink2"))
```

###Standardized

```{r}
hdgii_scaled <- scale(hdgii)
pca_scaled <- prcomp(hdgii_scaled)
summary(pca_scaled)
```

```{r}
biplot(pca_scaled, choices = 1:2, cex = c(0.8, 1), col = c("grey40", "deeppink2"))
```


##Multiple Correspondence Analysis (MCA) with the tea dataset

```{r}
library(FactoMineR)
data("tea")
str(tea)
```

```{r}
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")

tea_time <- select(tea, one_of(keep_columns))


summary(tea_time)
str(tea_time)

```

```{r}

```

